---
title: "Understanding and Predicting Recidivism"
author: "Jeff Westerhof"
date: '2022-06-21'
output:
  html_document: default
  pdf_document: default
---



Import & clean the data

```{r}
rec_raw <- read.csv("C:/Users/jwest/Documents/Final Project/NIJ_s_Recidivism_Challenge_Full_Dataset.csv", header = T, stringsAsFactors = TRUE)
  
summary(rec_raw)
head(rec_raw)

#removing rows with more than 5 missing values
#removing rows with more than 4 removes 20% of dataset,threshold increased to 5
#and only 243 rows removed
rec <- rec_raw[rowSums(is.na(rec_raw)) < 6, ]
summary(rec)

#Imputing Supervision Level from Risk Score
sum(rec$Supervision_Level_First == "")
library(epiDisplay)
risk<-tabpct(rec$Supervision_Risk_Score_First,
             rec$Supervision_Level_First)
rec$Supervision_Level_First[rec$Supervision_Level_First == "" & rec$Supervision_Risk_Score_First<6] = 'Standard'
rec$Supervision_Level_First[rec$Supervision_Level_First == "" & rec$Supervision_Risk_Score_First>5 & rec$Supervision_Risk_Score_First < 9]= 'High'
rec$Supervision_Level_First[rec$Supervision_Level_First == "" & rec$Supervision_Risk_Score_First>8] = 'Specialized'
rec$Supervision_Level_First[rec$Supervision_Level_First == "" & is.na(rec$Supervision_Risk_Score_First)] = 'Standard'

#Imputing Gang Affiliation by mode
sum(rec$Gang_Affiliated == "")
which.max(table(rec$Gang_Affiliated))
rec$Gang_Affiliated[rec$Gang_Affiliated == ""] = 'false'
summary(rec)
#Imputing Gender by mode
sum(rec$Gender == "")
which.max(table(rec$Gender))
rec$Gender[rec$Gender == ""] = 'M'

#Imputing Offence: category other
sum(rec$Prison_Offense== '')
rec$Prison_Offense[rec$Prison_Offense== '']= 'Other'

summary(rec)
#Imputing Drug tests. Because all the NAs for the drug tests are the same records,
#going on the assumption that drug testing was not part of their release conditions
#Since the median is 0, will assign 0 to these records and not the mean
sum(is.na(rec$Avg_Days_per_DrugTest))
sum(is.na(rec$DrugTests_THC_Positive))
sum(is.na(rec$DrugTests_Cocaine_Positive))
sum(is.na(rec$DrugTests_Meth_Positive))
sum(is.na(rec$DrugTests_Other_Positive)& is.na(rec$DrugTests_Meth_Positive)
    & is.na(rec$DrugTests_Cocaine_Positive)& is.na(rec$DrugTests_THC_Positive))
rec$Avg_Days_per_DrugTest[is.na(rec$Avg_Days_per_DrugTest) &
                            is.na(rec$DrugTests_Cocaine_Positive)]= 0
rec$DrugTests_THC_Positive[is.na(rec$DrugTests_THC_Positive)] = 0
rec$DrugTests_Cocaine_Positive[is.na(rec$DrugTests_Cocaine_Positive)]= 0
rec$DrugTests_Meth_Positive[is.na(rec$DrugTests_Meth_Positive)]=0
rec$DrugTests_Other_Positive[is.na(rec$DrugTests_Other_Positive)] = 0

#remaining NAs for Avg_Days_per_DrugTest assigned the mean
summary(rec$Avg_Days_per_DrugTest)
rec$Avg_Days_per_DrugTest[is.na(rec$Avg_Days_per_DrugTest)]= 75.12

#Imputing Employment Data
sum(is.na(rec$Percent_Days_Employed))
sum(is.na(rec$Jobs_Per_Year))
summary(rec$Percent_Days_Employed)
summary(rec$Jobs_Per_Year)
rec$Percent_Days_Employed[is.na(rec$Percent_Days_Employed)]= .4845
rec$Jobs_Per_Year[is.na(rec$Jobs_Per_Year)] = .7722
str(rec)
```

Some descriptive statistics

```{r}
summary(rec)
str(rec)
install.packages('epiDisplay')
library(epiDisplay)
race <-tabpct(rec$Race, rec$Recidivism_Within_3years, percent = 'row',
              main = 'Recidivism & Race', xlab = 'Race', ylab = 'Recidivism',
              col = c('green','red'))
age<-tabpct(rec$Age_at_Release, rec$Recidivism_Within_3years, percent = 'row',
            main = 'Recidivism & Age', xlab = 'Age', ylab = 'Recidivism',
            col = c('green','red'))

gender<-tabpct(rec$Gender, rec$Recidivism_Within_3years, percent = 'row',
               main = 'Recidivism & Gender', xlab = 'Gender', ylab = 'Recidivism',
               col = c('green','red'))

education<-tabpct(rec$Education_Level, rec$Recidivism_Within_3years, percent = 'row',
                  main = 'Recidivism & Education', xlab = 'Education', ylab = 'Recidivism',
                  col = c('green','red'))
dependents<-tabpct(rec$Dependents, rec$Recidivism_Within_3years, percent = 'row',
                   main = 'Recidivism & Dependents', xlab = 'Dependents', ylab = 'Recidivism',
                   col = c('green','red'))
offence<-tabpct(rec$Prison_Offense, rec$Recidivism_Within_3years, percent = 'row',
                main = 'Recidivism & Offence', xlab = 'Offence', ylab = 'Recidivism',
                col = c('green','red'))
sentence<-tabpct(rec$Prison_Years, rec$Recidivism_Within_3years, percent = 'row',
                 main = 'Recidivism & Sentence', xlab = 'Sentence', ylab = 'Recidivism',
                 col = c('green','red'))
felony<-tabpct(rec$Prior_Arrest_Episodes_Felony, rec$Recidivism_Within_3years, percent = 'row',
               main = 'Recidivism & Felony Arrests', xlab = 'Arrests', ylab = 'Recidivism',
               col = c('green','red'))
Misdemeanor<-tabpct(rec$Prior_Arrest_Episodes_Misd, rec$Recidivism_Within_3years, percent = 'row',
                    main = 'Recidivism & Misdemeanor Arrests', xlab = 'Arrests', ylab = 'Recidivism',
                    col = c('green','red'))
MH_condition<-tabpct(rec$Condition_MH_SA, rec$Recidivism_Within_3years, percent = 'row',
                     main = 'Recidivism & Mental health/Substance', xlab = 'MH conditions?', ylab = 'Recidivism',
                     col = c('green','red'))
Cog_condition<-tabpct(rec$Condition_Cog_Ed, rec$Recidivism_Within_3years, percent = 'row',
                      main = 'Recidivism & Cognitive conditions', xlab = 'Cog conditions?', ylab = 'Recidivism',
                      col = c('green','red'))
Oth_condition<-tabpct(rec$Condition_Othe, rec$Recidivism_Within_3years, percent = 'row',
                      main = 'Recidivism & Other conditions', xlab = 'Oth conditions?', ylab = 'Recidivism',
                      col = c('green','red'))
delinquency<-tabpct(rec$Delinquency_Reports, rec$Recidivism_Within_3years, percent = 'row',
                    main = 'Recidivism & delinquency', xlab = 'Delinquencies', ylab = 'Recidivism',
                    col = c('green','red'))
attendance<-tabpct(rec$Program_Attendances, rec$Recidivism_Within_3years, percent = 'row',
                   main = 'Recidivism & Program attendance', xlab = 'Attendances', ylab = 'Recidivism',
                   col = c('green','red'))
absences<-tabpct(rec$Program_UnexcusedAbsences, rec$Recidivism_Within_3years, percent = 'row',
                 main = 'Recidivism & Program absence', xlab = 'Absence', ylab = 'Recidivism',
                 col = c('green','red'))
residence<-tabpct(rec$Residence_Changes, rec$Recidivism_Within_3years, percent = 'row',
                  main = 'Recidivism & Residence changes', xlab = 'Residences', ylab = 'Recidivism',
                  col = c('green','red'))

sapply(rec[,c(2,3,4,5,6,8,9,10,11,12,13,14,31,32,33,38,39,40,41,50)], function(x) table(x)/nrow(rec))


#Quantitative Variables
sd(rec$Avg_Days_per_DrugTest[!is.na(rec$Avg_Days_per_DrugTest)])
sd(rec$DrugTests_THC_Positive[!is.na(rec$DrugTests_THC_Positive)])
sd(rec$DrugTests_Cocaine_Positive[!is.na(rec$DrugTests_Cocaine_Positive)])
sd(rec$DrugTests_Meth_Positive[!is.na(rec$DrugTests_Meth_Positive)])
sd(rec$DrugTests_Other_Positive[!is.na(rec$DrugTests_Other_Positive)])
sd(rec$Percent_Days_Employed[!is.na(rec$Percent_Days_Employed)])
sd(rec$Jobs_Per_Year[!is.na(rec$Jobs_Per_Year)])


hist(rec$Supervision_Risk_Score_First)
hist(rec$Percent_Days_Employed)
hist(rec$Jobs_Per_Year)
hist(rec$DrugTests_THC_Positive)
hist(rec$Avg_Days_per_DrugTest)
hist(rec$Avg_Days_per_DrugTest)

cor(rec[,c(42:48)])

```

Partition data to: Demographic & prior criminal history factors, demographic
& post release factors, All data
```{r}
prior_dem<-rec[,c(2,4:6,9:30,50)]

all_data<- rec[,c(2,4:6,9:50)]

post_dem<- rec[,c(2,4:6,9:10,31:50)]
```

Random Forest for feature importance & classification
```{r}
install.packages("randomForest", repos = "http://cran.us.r-project.org")

library(randomForest)
rf_prior<-randomForest(Recidivism_Within_3years~.,data = prior_dem, 
                 importance= TRUE, ntree = 100)
rf_all<-randomForest(Recidivism_Within_3years~.,data = all_data, 
                         importance= TRUE, ntree = 100)
importance(rf_prior)
importance(rf_all)
varImpPlot(rf_prior)
varImpPlot(rf_all)

#Classification

install.packages("caret", repos = "http://cran.us.r-project.org")
library(caret)
set.seed(222)
summary(prior_dem)
rf_prior_index<- createDataPartition(prior_dem[,"Recidivism_Within_3years"],
                                     p = 0.7, list = FALSE)
rfprior_train <-prior_dem[rf_prior_index,]
rfprior_test <- prior_dem[-rf_prior_index,]

prior_control <- trainControl(method = "cv", number = 10)


summary(rf_prior)
rf_prior_pred <- predict(rf_prior, rfprior_test)
confusionMatrix(rfprior_test[,"Recidivism_Within_3years"],rf_prior_pred,
                positive = "true")

rf_all_index<- createDataPartition(all_data[,"Recidivism_Within_3years"],
                                   p = 0.7, list = FALSE)
rfall_train <-all_data[rf_all_index,]
rfall_test <- all_data[-rf_all_index,]

prior_control <- trainControl(method = "cv", number = 10)


print(rf_all)
summary(rf_all)
rf_all$results
rf_all_pred <- predict(rf_all, rfall_test)
confusionMatrix(rfall_test[,"Recidivism_Within_3years"],rf_all_pred,
                positive = "true")

```
Classification Using knn
```{r}
knn_data_prior<-prior_dem
knn_data_all <-all_data
knn_data_post <-post_dem
summary(knn_data_all)


knn_data_prior[1:26]= sapply(knn_data_prior[1:26], unclass)
normalize <- function(x){(x - min(x)) / (max(x) - min(x))}
knn_data_prior[1:26]=sapply(knn_data_prior[1:26], normalize)
summary(knn_data_prior)

knn_data_all[1:45]= sapply(knn_data_all[1:45], unclass)
knn_data_all[1:45]= sapply(knn_data_all[1:45], normalize)
summary(knn_data_all)

knn_data_post[1:25]= sapply(knn_data_post[1:25], unclass)
knn_data_post[1:25]=sapply(knn_data_post[1:25], normalize)
summary(knn_data_post)

set.seed(222)

knn_prior_index<- createDataPartition(knn_data_prior[,"Recidivism_Within_3years"],
                                      p = 0.7, list = FALSE)
knn_prior_train <-knn_data_prior[knn_prior_index,]
knn_prior_test <-knn_data_prior[-knn_prior_index,]

knn_prior_control <- trainControl(method = "cv", number = 10)

knn_prior <- train(Recidivism_Within_3years~.,data = knn_prior_train,
                   method = "knn", trControl = knn_prior_control, 
                   tuneGrid =data.frame(k=seq(5,100,by=10)))
print(knn_prior)
summary(knn_prior)
knn_prior$results
knn_prior_pred <- predict(knn_prior, knn_prior_test)
confusionMatrix(knn_prior_test[,"Recidivism_Within_3years"],knn_prior_pred,
                positive = 'true')

###KNN USING ALL DATA
knn_all_index<- createDataPartition(knn_data_all[,"Recidivism_Within_3years"],
                                    p = 0.7, list = FALSE)
knn_all_train <-knn_data_all[knn_all_index,]
knn_all_test <-knn_data_all[-knn_all_index,]

knn_all_control <- trainControl(method = "cv", number = 10)

knn_all <- train(Recidivism_Within_3years~.,data = knn_all_train,
                 method = "knn", trControl = knn_all_control, 
                 tuneGrid =data.frame(k=seq(5,100,by=10)))
print(knn_all)
summary(knn_all)
knn_all$results
knn_all_pred <- predict(knn_all, knn_all_test)
confusionMatrix(knn_all_test[,"Recidivism_Within_3years"],knn_all_pred,
                positive = 'true')
plot(knn_all)


##USING POST DATA
knn_post_index<- createDataPartition(knn_data_post[,"Recidivism_Within_3years"],
                                     p = 0.7, list = FALSE)
knn_post_train <-knn_data_post[knn_post_index,]
knn_post_test <-knn_data_post[-knn_post_index,]

knn_post_control <- trainControl(method = "cv", number = 10)

knn_post <- train(Recidivism_Within_3years~.,data = knn_post_train,
                  method = "knn", trControl = knn_post_control, 
                  tuneGrid =data.frame(k=seq(5,100,by=10)))
print(knn_post)
summary(knn_post)
knn_post$results
knn_post_pred <- predict(knn_post, knn_post_test)
confusionMatrix(knn_post_test[,"Recidivism_Within_3years"],knn_post_pred,
                positive = 'true')
```

Using Logistic Regression to generate risk scores based on pre-release factors and pre- and post-release factors. Accuracy as a classification method also tested. Comparing post release factors between the group whose risk scores increased and the group whose scores decreased
```{r}
index <- sample(1:nrow(prior_dem), .7 * nrow(prior_dem))
lr_prior_train <- prior_dem[index,]
lr_prior_test <- prior_dem[-index,]

lr_prior<-glm(Recidivism_Within_3years ~.,lr_prior_train, family = 'binomial')

lr_prior_pred <- predict(lr_prior, lr_prior_test, type="response")
lr_prior_class <- ifelse(lr_prior_pred>=0.5, "true", "false")
lr_prior_confmatrix <- table(actual = lr_prior_test$Recidivism_Within_3years,
                             predicted = lr_prior_class)
lr_prior_confmatrix

all_data_risk <- all_data
index <- sample(1:nrow(all_data), .7 * nrow(all_data))
lr_all_train <- all_data[index,]
lr_all_test <- all_data[-index,]



lr_all<-glm(Recidivism_Within_3years ~.,lr_all_train, family = 'binomial')
summary(lr_all)

lr_all_pred <- predict(lr_all, lr_all_test, type="response")
lr_all_class <- ifelse(lr_all_pred>=0.5, "true", "false")
lr_all_confmatrix <- table(actual = lr_all_test$Recidivism_Within_3years,
                           predicted = lr_all_class)
lr_all_confmatrix

all_data_risk$prior_prob = predict(lr_prior, all_data_risk, type = "response")
all_data_risk$all_prob = predict(lr_all, all_data_risk, type = "response")
hist(all_data_risk$all_prob)
all_data_risk$prob_dif = all_data_risk$all_prob - all_data_risk$prior_prob
hist(all_data_risk$prob_dif)
sd(all_data_risk$prior_prob)
all_data_risk$change = "Higher"
all_data_risk$change[all_data_risk$prob_dif<0] = "Lower"
lower_prob <- all_data_risk[all_data_risk$prob_dif < -0.1,]
higher_prob <- all_data_risk[all_data_risk$prob_dif > 0.1,]

hist(higher_prob$Jobs_Per_Year)
hist(lower_prob$Jobs_Per_Year)
hist(higher_prob$Percent_Days_Employed)
hist(lower_prob$Percent_Days_Employed)
plot(lower_prob$Recidivism_Within_3years)

plot(higher_prob$Recidivism_Within_3years)
wilcox.test(higher_prob$Percent_Days_Employed,
            lower_prob$Percent_Days_Employed, exact= F)
wilcox.test(higher_prob$Jobs_Per_Year,
            lower_prob$Jobs_Per_Year, exact= F)
resid <-table(all_data_risk$change,all_data_risk$Residence_Changes)
resid
chisq.test(resid)

```

